### 2021-07-04 rad-icfp-2021/appendix1.tex
redoing format appendix proof

\MH{change to make it nice. be more explicit about optimisations. 0 should be ZerosLike. 
Also, I am freely referring to the variables from the context or bound in theoriginal term, should be clarified.}

@@

 \begin{proof}
\begin{align*}
    \nabla_A \text{prod(A)} 
    &= \nabla_A(\text{reduce} * 1 A) \\
    &\defeq  \Big(\lambda A',z.
    \begin{aligned}
        & let A_1 = scanl * 1 A in \\
        & let A_2 = map2 (x,y -> y) A_1 A in\\
        & let A_3 = shift1 (map2 (x,y -> x) A_1 A) in\\
        & let A_4 = scanr * 1 A_2 in \\
        & map3 (a,b,c -> a+b*c*z) A' A_4 A_3
    \end{aligned}
    \Big)(0,1)\\
    &\stackrel{\beta-reduction}{=}  \begin{aligned}
        & let A_1 = scanl * 1 A in &\\
        & let A_2 = map2 (x,y -> y) A_1 A in & A_2=  A\\
        & let A_3 = shift1 (map2 (x,y -> x) A_1 A) in & A_3 = shift1 A1\\
        & let A_4 = scanr * 1 A_2 in & \\
        & map3 (a,b,c -> a+b*c*1) 0 A_4 A_3 & map2 * A_4 A_3
    \end{aligned}\\
    &=  \begin{aligned}
        & let A_1 = scanl * 1 A in & \\
        & let A_3 = shift1 A1 in &\\
        & let A_4 = scanr * 1 A in &\text{forward substitution} A_2\\
        & map2 * A_4 A_3
    \end{aligned}\\
    &\stackrel{\eta-reduction}{=} map2 * (scanr x 1 A) (shift1(scanl * 1 A))
\end{align*}
 \end{proof}

@@

\begin{align*}
    \nabla_A sum(A) 
    &= \nabla_A reduce + 0 A \\
    & \defeq \Big(\lambda A',z.
    \begin{aligned}
        & let A_1 = scanl * 1 A in \\
        & let A_2 = map2 (x,y -> 1) A_1 A in \\
        & let A_3 = shift1 (map2 (x,y -> 1) A_1 A) in \\
        & let A_4 = scanr * 1 A_2 in \\
        & map3 (a,b,c -> a+b*c*z) A' A_4 A_3
    \end{aligned}
    \Big)(0,1)\\
    &{\beta-reduction}{=}
    \begin{aligned}
        & let A_1 = scanl * 1 A in &\\
        & let A_2 = map2 (x,y -> y) A_1 A in & A_2= OnesLike(A) \\
        & let A_3 = shift1 (map2 (x,y -> x) A_1 A) in & A_3 = shift1 OnesLike(A)\\
        & let A_4 = scanr * 1 A_2 in & \\
        & map3 (a,b,c -> a+b*c*1) 0 A_4 A_3 & map2 * A_4 A_3
    \end{aligned}\\
    &= \begin{aligned}
        & let A_1 = scanl * 1 A in & \\
        & let A_2= OnesLike(A) in &  \\
        & let A_3 = shift1 OnesLike(A) in & A_3 = OnesLike(A) \\
        & let A_4 = scanr * 1 A_2 in & A_4 = OnesLike(A_2) \\
        & map2 * A_4 A_3 &
    \end{aligned}\\
    &= map2 * OnesLike(A) OnesLike(A) \\
    &= OnesLike(A)
\end{align*}
 \end{proof}

@@

 \begin{proof}
The reverse derivative of !map2 * A B! is given by 
\begin{align*}
    & \lambda A',B',C'. \\
    & let C_1 =map2 (a,b -> b) A B in \\
    & let C_2 =map2 (a,b -> a) A B in \\
    & (map3 (a,b,c -> a+b*c) A' C_1 C', map3 (a,b,c -> a+b*c) A' C_2 C')
\end{align*}

@@

Then the gradient of !dot(A,B)! is given by
\begin{align*}
    \nabla dot(A,B) 
    &= \nabla let C = map (x -> 1) A in map2 * B C\\
    &\defeq \lambda A',B',C',z. 
    \left.\begin{aligned}
        & let A_1 = scanl * 1 A in \\
        & let A_2 = map2 (x,y -> 1) A_1 A in \\
        & let A_3 = shift1 (map2 (x,y -> 1) A_1 A) in \\
        & let A_4 = scanr * 1 A_2 in \\
        & Y(A', B', map3 (a,b,c -> a+b*c*z) A' A_4 A_3)
    \end{aligned}\right\}same reduction as previously
    \\
    &\quad(0,0,0,1) \\
    &= Y(0,0,OnesLike(C)) \\
    &\defeq 
    \begin{aligned}
        & let C_1 =map2 (a,b -> b) A B in & C_1 = B\\
        & let C_2 =map2 (a,b -> a) A B in & C_2 = A\\
        & (map3 (a,b,c -> a+b*c) 0 C_1 OnesLike(C), & map2 & C_1 OnesLike(C)
        & map3 (a,b,c -> a+b*c) 0 C_2 OnesLike(C)) & map2 & C_2 OnesLike(C)
    \end{aligned}\\
    &= \begin{aligned}
        & let C_1 = B & \text{forward substitution}\\
        & let C_2 = A \text{forward substitution}\\
        & map2 & C_1 OnesLike(C) & C_1
        & map2 & C_2 OnesLike(C) & C_2
    \end{aligned}\\
    &= (B, A)
\end{align*}
So if we are only interested in the gradient w.r.t. A, this indeed gives us B.
 \end{proof}

>>>>

\MH{be more explicit about optimisations. 0 should be ZerosLike. A bit outdated compared to current D transformation.
Also, I am freely referring to the variables from the context or bound in the original term, should be clarified.}

@@

 \begin{proof}
The gradient of  derivative $\nabla_A$!prod(A)! is given by

\begin{tabular}{c l r}
    & $\nabla_A$ !prod(A)! & \\
    =&  $\nabla_A$(!reduce * 1 A!) & \\
    $\defeq$ & \Big($\lambda$ !A',z. let A1 = scanl * 1 A in! & \\
    & !let A2 = map2 (x,y -> y) A1 A in! & \\ 
    & !let A3 = shift1 (map2 (x,y -> x) A1 A) in! & \\
    & !let A4 = scanr * 1 A2 in! &\\
    & !map3 (a,b,c -> a+b*c*z) A' A4 A3! & \\
    & \Big)(0,1)\\
     $\stackrel{\beta-reduction}{=}$  & !let A1 = scanl * 1 A in! & \\
    & !let A2 = map2 (x,y -> y) A1 A in! & !A2=A! \\
    & !let A3 = shift1 (map2 (x,y -> x) A1 A) in! & !A3=shift1 A1!\\
    & !let A4 = scanr * 1 A2 in! & \\
    & !map3 (a,b,c -> a+b*c*1) 0 A4 A3! & !map2 * A4 A3! \\
    = & !let A1 = scanl * 1 A in! & \\
    & !let A3 = shift1 A1 in! &\\
    & !let A4 = scanr * 1 A in! & forward substitution !A2!\\
    & !map2 * A4 A3! & \\
    $\stackrel{\eta-reduction}{=}$ & !map2 * (scanr x 1 A) (shift1(scanl * 1 A))! &
\end{tabular}
 \end{proof}

@@

 \begin{proof}
The gradient of  derivative $\nabla_A$!sum(A)! is given by

    \begin{tabular}{c l r}
    & $\nabla_A$ !sum(A)! & \\
    = & $\nabla_A$ !reduce + 0 A! & \\
    $\defeq$ & \Big($\lambda$ !A',z. let A1 = scanl * 1 A in! \\
    & !let A2 = map2 (x,y -> 1) A1 A in! & \\
    & !let A3 = shift1 (map2 (x,y -> 1) A1 A) in! & \\
    & !let A4 = scanr * 1 A2 in! & \\
    & !map3 (a,b,c -> a+b*c*z) A' A4 A3! &
    &  \Big)(0,1) & \\
    $\stackrel{\beta-reduction}{=}$  & !let A1 = scanl * 1 A in! & \\
    & !let A2 = map2 (x,y -> y) A1 A in! & !A2=OnesLike(A)! \\
    & !let A3 = shift1 (map2 (x,y -> x) A1 A) in! & !A3=shift1 OnesLike(A)!\\
    & !let A4 = scanr * 1 A2 in! & \\
    & !map3 (a,b,c -> a+b*c*1) 0 A4 A3! & !map2 * A4 A3! & \\
    = & !let A1 = scanl * 1 A in! & \\
    & !let A2= OnesLike(A) in! &  \\
    & !let A3 = shift1 OnesLike(A) in! & !A3=OnesLike(A)! \\
    & !let A4 = scanr * 1 A2 in! & !A4=OnesLike(A2)! \\
    & !map2 * A4 A3! & \\
    = & !map2 * OnesLike(A) OnesLike(A)! & \\
    = & !OnesLike(A)!
    \end{tabular}
\end{align*}
 \end{proof}

@@

 \begin{proof}
The reverse derivative of !map2 * A B! is given by

\begin{tabular}{c l}
    & $\lambda$ !A',B',C'.! \\
    & !let C1 =map2 (a,b -> b) A B in! \\
    & !let C2 =map2 (a,b -> a) A B in! \\
    & !(map3 (a,b,c -> a+b*c) A' C1 C', map3 (a,b,c -> a+b*c) A' C2 C')!
\end{tabular}

@@

Then the gradient of !dot(A,B)! is given by

\begin{tabular}{c l r}
    & $\nabla$ !dot(A,B)! & \\
    = & $\nabla$ !let C = map (x -> 1) A in map2 * B C! & \\
    $\defeq$ & $\Big(\lambda$ A',B',C',z. & \\
    & !let A1 = scanl * 1 A in!  \hspace{4cm}\rdelim\}{5}{3mm}[\parbox{40mm}{same reduction\\as previously}]\\
    & !let A1 = scanl * 1 A in! & \\
    & !let A2 = map2 (x,y -> 1) A1 A in! & \\
    & !let A3 = shift1 (map2 (x,y -> 1) A1 A) in! &\\
    & !let A4 = scanr * 1 A2 in! &\\
    & !Y(A', B', map3 (a,b,c -> a+b*c*z) A' A4 A3)!\Big) &  \\
    & \quad !(0,0,0,1)! & \\
    = & !Y(0,0,OnesLike(C))! & \\
    $\defeq$ & !let C1=map2 (a,b -> b) A B in! & !C1 = B! \\
    & !let C2=map2 (a,b -> a) A B in! & !C2 = A! \\
    & !(map3 (a,b,c -> a+b*c) 0 C1 OnesLike(C)!, & !map2 C1 OnesLike(C)! \\
    & !map3 (a,b,c -> a+b*c) 0 C2 OnesLike(C))!  & !map2 C2 OnesLike(C)! \\
    = & !let C1=B! & forward substitution \\
    & !let C2=A! & forward substitution \\
    & !map2 C1 OnesLike(C)! & !C1! \\
    & !map2 C2 OnesLike(C)! & !C2! \\
    = & !(B, A)!
\end{tabular}

So if we are only interested in the gradient w.r.t. A, this indeed gives us B.
\end{proof}

### 2021-07-04 rad-icfp-2021/purerad.tex
redoing format appendix proof

For the sake of simplicity in the presentation, the bound function in !reduce! 
is restricted to having no free variables. For the same reason, we currently restrict to arrays of reals.
We show how to lift these restrictions in Section~\ref{sec:generalisation}.

@@

TODO:redo
Similarly, the target language contains more array primitives which are used to define the reverse derivatives of array operations. 
!scanl! is a !foldl! that stores all intermediate results. 
!scanr! is similar, but reads the array from right to left, and also stores from right to left.
!reduce! is a !foldl! for which the function is associative. 
This means in practice that it can be computed faster.
Finally, we add a somewhat strange operator which we call !shift1!. 
It takes an array of size $n$,  shifts all the element of the array by one, 
forgetting the last element, and puts a one in the first place. 
This operator naturally shows up when differentiating !reduce!.

The types and terms are presented in Figure~\ref{fig:target_grammar}.
We are not interested here with partial applications and our language is pure, so functions take $n$ arguments.
Lambda abstractions and applications will be removed during partial evaluation and this notation makes reading slightly easier.

@@

The type system for the extended grammar of the target language is presented in Figure~\ref{fig:target_typesystem}.

@@

TODO:redo
In Figure~\ref{fig:direct_diff_macro} we present the direct transformation for pure reverse mode differentiation.
We explain in the next Section (\ref{sub:Partial evaluation and optimisation}) how to optimise it further and compute the gradient.
This transformation may be hard to read and prove correct. 
In Section~\ref{sec:unf} we decompose this transformation into several easier 
and more intuitive steps via a new intermediate representation which we call UNF.

TODO: on the examples

@@

TODO: on the examples

@@

Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite
!fun (x1,...,xn) -> (fun (y1,...,yn) -> (f1,...,fn))(e1,...,en)! 
$\transto$ 
!fun (x1,...,xn) -> let y1,...,yn = e1,...,en in (f1,...,fn)!
$\transto$
!fun (x1,...,xn) -> (f1[e1/y1],...,fn[en/yn])!
Now, importantly, every !yi! is used at most once (except for !map2! again and !reduce! which require a more detailed analysis)
which garantees that this will not blow up the complexity.

>>>>

!reduce! is a fold left operator for which the function is assumed to be associative, 
and the provided initial value should be a unit of the binary operation.
It is a well-known parallel friendly construct. For the sake of simplicity in the presentation, the bound function in !reduce! 
is restricted to having no free variables. For the same reason, we currently restrict to arrays of reals.
We show how to lift these restrictions in Section~\ref{sec:generalisation}.

@@

Similarly, the target language contains more array primitives which are used to define the reverse derivatives of array operations. 
Scan left !scanl! is similar to fold left but also stores all the intermediate results in a array, which it returns.
In the same vein, scan right !scanr! perform a fold left by reading the array from right to left and also storing 
the intermediate results in a array from right to left.

Finally, we add a new shift operator !shift1!. 
It takes an array of size $n$,  shifts all the element of the array by one, 
forgetting the last element, and puts a one in the first place. 
This somewhat ad-hoc operator naturally shows up when differentiating fold-like operators.

The types and terms are presented in Figure~\ref{fig:target_grammar}.
Our lambda abstractions take $n$ arguments as we are not interested here with partial applications. 
In fact, the lambda abstractions introduced by reverse-mode will be removed during partial evaluation 
and the notation with lambda abstractions having $n$ bound variables makes reading slightly easier.
We note that we don't actually need the full power of higher-order because we only use lambda abstractions over variables of ground types
and let expressions binding such lambda abstractions. So we only require the target language to be second-order.
The type system for the extended grammar of the target language is presented in Figure~\ref{fig:target_typesystem}.

@@

@@

TODO:redo
In Figure~\ref{fig:direct_diff_macro} we present our direct transformation from the source language to the target language for pure reverse mode differentiation.
We explain in the next Section (\ref{sub:Partial evaluation and optimisation}) how to optimise it further and compute the gradient.
Admittedly, this transformation may be hard to read, and prove correct. 
In Section~\ref{sec:unf} we decompose this transformation into three simpler steps via a novel intermediate representation (UNF).

\begin{example}
    TODO
\end{example}

@@

\begin{example}
    TODO
\end{example}

@@

Using the invariant above and the inlining and forward substitution given in Figure~\ref{fig:optim}, this allows us to rewrite
!fun (x1,...,xn) -> (fun (y1,...,yn) -> (f1,...,fn))(e1,...,en)! 
$\transto$ 
!fun (x1,...,xn) -> let y1,...,yn = e1,...,en in (f1,...,fn)!
$\transto$
!fun (x1,...,xn) -> (f1[e1/y1],...,fn[en/yn])!

TODO: redo
Now, importantly, every !yi! is used at most once (except for !map2! again and !reduce! which require a more detailed analysis)
which garantees that this will not blow up the complexity.
